---
title: "Report"
author: "Keliang Xu, Shuting Li, Tao He"
date: "3/2/2022"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install.packages("foreign")
library(ggplot2)
library(openxlsx)
library(reshape2)
library(dplyr)
library(tidyr)
library("formattable")
library(stats)
```

# 1 Abstract
&nbsp; &nbsp;

This report consists of 4 main sections: Introduction, Checking and Testing, Discussion, and Conclusion.

# 2 Introduction

## 2.1 Data Set Introduction

&nbsp; &nbsp;The variables of interest are “Taxa”, “Site” and “Carbon and Nitrogen Level''. The “Taxa '' includes four types of animals, which are turkey, cottontails, deer, and hare, respectively. The “site” includes two cities, which are BAOX and PATT. 

&nbsp; &nbsp;The “Carbon and Nitrogen level” includes stable carbon and nitrogen isotopes from two closely related sites (BAOX and PATT) from the same area and period. These enter bones by eating and reflect the isotopic values of food sources. Carbon isotopes tell us about the level of human-grown plants in the diet, such as maize (corn). Carbon isotopes from collagen are protein-derived. Carbon isotopes from apatite (carbonate) reflect the whole diet. Nitrogen isotopes tell us about the protein in the diet.



## 2.2 Client’s Work


&nbsp; &nbsp;So far, the client has:
  
&nbsp; &nbsp;1.) Use ANOVA to test for the differences between BAOX and PATT for each species.
  
&nbsp; &nbsp;2.) Examine box-plots and test just those groupings that look different.


## 2.3 Client Requests and Deliverables

&nbsp; &nbsp;1.) Check all her works, including the summary data table and box plots.

&nbsp; &nbsp;2.) Use statistical methods to show there is no significant difference in specific animals’ diets from BAOX and PATT, except cottontails.


# 3 Checking and Testing

## 3.1 Checking 

```{r, include=FALSE}
#read data
data<-read.xlsx("for_stats_peeps.xlsx")
data<-data[,-2:-6]
data<-data[,-4]
removedata<-c("MC381","MC389","MC393","MC76")
#data <- na.omit(data)
i<-1
for(i in i:4) data<-subset(data,data$ID!=removedata[i])

```

```{r, include=FALSE}
#tidy data
data2<-data[,2:8]
data2<-data2[,-3]
data2<-data2[,-5]
aggdata<-melt(data2,id=(c("Taxa","Site")))

```
Dropped 4 missing value, their IDs are MC381,MC389,MC393,MC76.

```{r,include=FALSE}
summ <- aggdata %>%                              
  group_by(Taxa,Site,variable) %>% 
  na.omit(aggdata) %>%
  summarize(min = min(value),
            median = median(value),
            mean = mean(value),
            max = max(value),
            sd = sd(value),
            count=n())
#formattable(summ)

```

## EDA
```{r,echo=FALSE,include=FALSE,warning=FALSE}
ggplot(data = aggdata, aes(x=variable,y=value, fill=variable)) + 
  geom_boxplot()+
  geom_jitter(shape=16, position=position_jitter(0.2))+
  facet_wrap(~Taxa,nrow = 1)
```

In this plot, we compared specific isotopes level in 4 taxon. It showed the level of d15N_air is less diverse than other 2 isotopes.
```{r, echo=FALSE,warning=FALSE}
ggplot(data = aggdata, aes(x=Taxa,y=value, fill=Taxa)) + 
  geom_boxplot()+
  geom_jitter(shape=16, position=position_jitter(0.2))+
  facet_wrap(~variable,nrow = 1)
```

Then we compared isotopes in two sites for each kind of animals. The plot shows that the median level of d13C_VPDB and d15N_air in BAOX is higher than Tlajinga of cottontail. Similarly, the median level of d13C_VPDB and d15N_air of hare also show difference in the boxplot.


```{r, echo=FALSE,warning=FALSE}
data_deer<-aggdata[which(aggdata$Taxa=="Deer"),]
data_Cottontail<-aggdata[which(aggdata$Taxa=="Cottontail"),]
data_Hare<-aggdata[which(aggdata$Taxa=="Hare"),]
data_Turkey<-aggdata[which(aggdata$Taxa=="Turkey"),]

deer1<-ggplot(data = data_deer, aes(x=variable,y=value, fill=variable)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Deer')+
  facet_wrap(~Site,nrow = 1)

deer2<-ggplot(data = data_deer, aes(x=Site,y=value, fill=Site)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Deer')+
  facet_wrap(~variable,nrow = 1)



cot1<-ggplot(data = data_Cottontail, aes(x=variable,y=value, fill=variable)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Cottontail')+
  facet_wrap(~Site,nrow = 1)

cot2<-ggplot(data = data_Cottontail, aes(x=Site,y=value, fill=Site)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Cottontail')+
  facet_wrap(~variable,nrow = 1)


hare1<-ggplot(data = data_Hare, aes(x=variable,y=value, fill=variable)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Hare')+
  facet_wrap(~Site,nrow = 1)

hare2<-ggplot(data = data_Hare, aes(x=Site,y=value, fill=Site)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Hare')+
  facet_wrap(~variable,nrow = 1)

turkey1<-ggplot(data = data_Turkey, aes(x=variable,y=value, fill=variable)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Turkey')+
  facet_wrap(~Site,nrow = 1)

turkey2<-ggplot(data = data_Turkey, aes(x=Site,y=value, fill=Site)) + 
  geom_boxplot()+
  #geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(title = 'Turkey')+
  facet_wrap(~variable,nrow = 1)

```

```{r, echo=FALSE,warning=FALSE}

#gridExtra::grid.arrange(deer1,cot1,nrow=2,ncol=1)
#gridExtra::grid.arrange(hare1,turkey1,nrow=2,ncol=1)

gridExtra::grid.arrange(deer2,cot2,nrow=2,ncol=1)
gridExtra::grid.arrange(hare2,turkey2,nrow=2,ncol=1)

```

### Check Outlier

```{r}
mean(filter(aggdata,Taxa=="Hare" & Site=="BAOX" & variable=="d15N_air")$value)
sd(filter(aggdata,Taxa=="Hare" & Site=="BAOX" & variable=="d15N_air")$value)
```
outlier falls on mean +- 2sd, we believe it is not a outlier.


## 3.2 Testing
## Density Test

We would like to use the T-tests to find the difference of diets between two sites.

Firstly, we checked the normality.

```{r,echo=FALSE,warning=FALSE}
# check the d13C_VPDB
aggdata1 <- filter(aggdata, variable== "d13C_VPDB" & Site == "BAOX")
ggplot1 <- ggplot(aggdata1, aes(value, colour = Taxa,fill=Taxa)) + geom_density(alpha = 0.3) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  labs(x = "d13C_VPDB", y = "density", title = "4 kinds of animals'd13C_VPDB level distribution in BOAX")


aggdata2 <- filter(aggdata, variable== "d13C_VPDB" & Site == "Tlajinga")
ggplot2 <- ggplot(aggdata2, aes(value, colour = Taxa,fill=Taxa)) + geom_density(alpha = 0.3) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  labs(x = "d13C_VPDB", y = "density", title = "4 kinds of animals'd13C_VPDB level distribution in Tlajinga")

gridExtra::grid.arrange(ggplot1, ggplot2)
```

```{r,echo=FALSE,warning=FALSE}
# check the d15N_air
aggdata3 <- filter(aggdata, variable== "d15N_air" & Site == "BAOX")
ggplot3 <- ggplot(aggdata3, aes(value, colour = Taxa,fill=Taxa)) + geom_density(alpha = 0.3) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  labs(x = "d13C_VPDB", y = "density", title = "4 kinds of animals'd15N_air level distribution in BOAX")


aggdata4 <- filter(aggdata, variable== "d15N_air" & Site == "Tlajinga")
ggplot4 <- ggplot(aggdata4, aes(value, colour = Taxa,fill=Taxa)) + geom_density(alpha = 0.3) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  labs(x = "d13C_VPDB", y = "density", title = "4 kinds of animals'd15N_air level distribution in Tlajinga")

gridExtra::grid.arrange(ggplot3, ggplot4)
```


```{r,echo=FALSE,warning=FALSE}
# check the ap_13C_VPDB
aggdata5 <- filter(aggdata, variable== "ap_13C_VPDB" & Site == "BAOX")
ggplot5 <- ggplot(aggdata5, aes(value, colour = Taxa,fill=Taxa)) + geom_density(alpha = 0.3) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  labs(x = "d13C_VPDB", y = "density", title = "4 kinds of animals'ap_13C_VPDB level distribution in BOAX")


aggdata6 <- filter(aggdata, variable== "ap_13C_VPDB" & Site == "Tlajinga")
ggplot6 <- ggplot(aggdata6, aes(value, colour = Taxa,fill=Taxa)) + geom_density(alpha = 0.3) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  labs(x = "d13C_VPDB", y = "density", title = "4 kinds of animals'ap_13C_VPDB level distribution in Tlajinga")

gridExtra::grid.arrange(ggplot5, ggplot6)
```

+ Normal Distribution Test (Q-Q Plot)

Take "Turkey" and "d15N_air" as an example:
```{r,echo=FALSE}
qqnorm(aggdata[aggdata$Taxa=="Turkey"&aggdata$variable=="d15N_air",]$value)
qqline(aggdata[aggdata$Taxa=="Turkey"&aggdata$variable=="d15N_air",]$value)
```

## Mann Whitney U Test

Mann-Whitney U Test for deer.

```{r,echo=FALSE,warning=FALSE}
data_deer_dc<-data_deer[which(data_deer$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_deer_dc)

data_deer_apc<-data_deer[which(data_deer$variable=="ap_13C_VPDB"),] %>% na.omit(data_deer)
wilcox.test(value~Site,data=data_deer_apc)

data_deer_n<-data_deer[which(data_deer$variable=="d15N_air"),]
wilcox.test(value~Site,data=data_deer_n)
```

Mann-Whitney U Test for Cottontail.

```{r,echo=FALSE,warning=FALSE}
data_cott_dc<-data_Cottontail[which(data_Cottontail$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_cott_dc)

data_cott_apc<-data_Cottontail[which(data_Cottontail$variable=="ap_13C_VPDB"),] 
wilcox.test(value~Site,data=data_cott_apc)

data_cott_n<-data_Cottontail[which(data_Cottontail$variable=="d15N_air"),]%>% na.omit(data_Cottontail)
wilcox.test(value~Site,data=data_cott_n)
```

Mann-Whitney U Test for Hare.

```{r,echo=FALSE,warning=FALSE}
data_hare_dc<-data_Hare[which(data_Hare$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_hare_dc)

data_hare_apc<-data_Hare[which(data_Hare$variable=="ap_13C_VPDB"),] 
wilcox.test(value~Site,data=data_hare_apc)

data_hare_n<-data_Hare[which(data_Hare$variable=="d15N_air"),]
wilcox.test(value~Site,data=data_hare_n)
```

Mann-Whitney U Test for Turkey.

```{r,echo=FALSE,warning=FALSE}
data_turkey_dc<-data_Turkey[which(data_Turkey$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_turkey_dc)

data_turkey_apc<-data_Turkey[which(data_Turkey$variable=="ap_13C_VPDB"),] 
wilcox.test(value~Site,data=data_turkey_apc)

data_turkey_n<-data_Turkey[which(data_Turkey$variable=="d15N_air"),]
wilcox.test(value~Site,data=data_turkey_n)
```

```{r}
#waining mean:
#  The impact of ties means the Wilcoxon rank sum distribution cannot be used to calculate exact p-values. If ties occur in our data and we have fewer than 50 observations, the wilcox.test function returns a normal approximated p-value along with a warning message that says “cannot compute exact p-value with ties”.


#Since this p-value is not less than 0.05, we fail to reject the null hypothesis. 


# Mann-Whitney U Test for Cottontail.

# W=52, p-value= 0.03792

# we can reject the null hypothesis:
# two sites animals are the same
```

## Check Mann-Whitney U Test with theory

```{r}
Utable<-read.xlsx("Mann-Whitney_Table_0.01.xlsx")

Utest<-function(set1,set2){
  l1<-length(set1)
  l2<-length(set2)
  set<-c(set1,set2)
  set<-rbind(set,c(rep(1,l1),rep(2,l2)))
  rset<-rank(set[1,])
  R1<-sum(rset[1:l1])
  R2<-sum(rset[(l1+1):(l1+l2)])
  U1<-l1*l2+l1*(l1+1)/2-R1
  U2<-l1*l2+l2*(l2+1)/2-R2
  print(min(U1,U2))
  print(Utable[l1,l2])
  min(U1,U2)<Utable[l1,l2]
}
```

Mann-Whitney U Test for deer.

```{r,echo=FALSE,warning=FALSE}
setA<-data_deer_dc[which(data_deer_dc$Site=="BAOX"),]
setB<-data_deer_dc[which(data_deer_dc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_deer_apc[which(data_deer_apc$Site=="BAOX"),]
setB<-data_deer_apc[which(data_deer_apc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_deer_n[which(data_deer_n$Site=="BAOX"),]
setB<-data_deer_n[which(data_deer_n$Site=="Tlajinga"),]
Utest(setA$value,setB$value)
```

Mann-Whitney U Test for Cottontail.

```{r,echo=FALSE,warning=FALSE}
setA<-data_cott_dc[which(data_cott_dc$Site=="BAOX"),]
setB<-data_cott_dc[which(data_cott_dc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_cott_apc[which(data_cott_apc$Site=="BAOX"),]
setB<-data_cott_apc[which(data_cott_apc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_cott_n[which(data_cott_n$Site=="BAOX"),]
setB<-data_cott_n[which(data_cott_n$Site=="Tlajinga"),]
Utest(setA$value,setB$value)
```

Mann-Whitney U Test for Hare.

```{r,echo=FALSE,warning=FALSE}
setA<-data_hare_dc[which(data_hare_dc$Site=="BAOX"),]
setB<-data_hare_dc[which(data_hare_dc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_hare_apc[which(data_hare_apc$Site=="BAOX"),]
setB<-data_hare_apc[which(data_hare_apc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_hare_n[which(data_hare_n$Site=="BAOX"),]
setB<-data_hare_n[which(data_hare_n$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

```

Mann-Whitney U Test for Turkey.

```{r,echo=FALSE,warning=FALSE}
setA<-data_turkey_dc[which(data_turkey_dc$Site=="BAOX"),]
setB<-data_turkey_dc[which(data_turkey_dc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_turkey_apc[which(data_turkey_apc$Site=="BAOX"),]
setB<-data_turkey_apc[which(data_turkey_apc$Site=="Tlajinga"),]
Utest(setA$value,setB$value)

setA<-data_turkey_n[which(data_turkey_n$Site=="BAOX"),]
setB<-data_turkey_n[which(data_turkey_n$Site=="Tlajinga"),]
Utest(setA$value,setB$value)
```


Because we find that Cottontail of d15N_air test reject the null hypothesis due to p-value.
We want to check Mann-Whitney U Test without R package again.
```{r,echo=FALSE,warning=FALSE}
mann <- subset(aggdata,select = -Site) %>%                         
  group_by(Taxa,variable) %>% 
  na.omit(aggdata) %>%
  summarize(rank = rank(value))
mann<- aggdata %>%                              
  group_by(Taxa,variable) %>% 
  na.omit(aggdata) %>%
  cbind(rank=mann$rank)
#formattable(mann)

mann_cott_n<-mann[which(mann$variable=="d15N_air" & mann$Taxa=="Cottontail"),]
setA<-mann_cott_n[mann_cott_n$Site=="BAOX",]
setB<-mann_cott_n[mann_cott_n$Site=="Tlajinga",]
n1<-length(setA$Taxa)
n2<-length(setB$Tlajinga)
R1<-sum(setA$rank)
R2<-sum(setB$rank)
U1<-n1*n2+n1*(n1+1)/2-R1
U2<-n1*n2+n2*(n2+1)/2-R2
#min(U1,U2)

```

 Alpha=0.01, test statistic(7) and our critical value(7) cannot reject

 Alpha=0.05. Since our test statistic (13) is greater than our critical value (7), we fail to reject the null hypothesis.


 Theory : https://www.statology.org/mann-whitney-u-test/
 Table : https://www.statology.org/mann-whitney-u-table/


## Next Step

We plan to try some statistic test or some regression to compare the isotopes level in two sites more precisely.

Two-Sample T-tests are as followings:
  
### "d13C_VPDB" level
  
```{r}
# Assumption
# H0 : v1  = v2
# Ha : v1 != v2
```

```{r}
# "d13C_VPDB" level in turkey groups
test.turkey.d13C <- filter(aggdata, Taxa == "Turkey" & variable== "d13C_VPDB")
with(test.turkey.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.4606
with(test.turkey.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.7027

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.turkey.d13C, var.equal = TRUE)
# p = 0.3323
# Not reject!
```

```{r}
# "d13C_VPDB" level in Cottontail groups
test.Cottontail.d13C <- filter(aggdata, Taxa == "Cottontail" & variable== "d13C_VPDB")
with(test.Cottontail.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.5969
with(test.Cottontail.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.3941

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Cottontail.d13C, var.equal = TRUE)
# p = 0.8162
# Not reject!
```

```{r}
# "d13C_VPDB" level in Deer groups
test.Deer.d13C <- filter(aggdata, Taxa == "Deer" & variable== "d13C_VPDB")
with(test.Deer.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.3914
with(test.Deer.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2177

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Deer.d13C, var.equal = TRUE)
# p = 0.6657
# Not reject!
```

```{r}
# "d13C_VPDB" level in Hare groups
test.Hare.d13C <- filter(aggdata, Taxa == "Hare" & variable== "d13C_VPDB")
with(test.Hare.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.2707
with(test.Hare.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2447

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.Hare.d13C, var.equal = TRUE)
# p = 0.3999
# Not reject!
```

### "d15N_air" level

```{r}
# "d15N_air" level in turkey groups
test.turkey.d15N <- filter(aggdata, Taxa == "Turkey" & variable== "d15N_air")
with(test.turkey.d15N, shapiro.test(value[Site == "BAOX"])) # p = 0.7717
with(test.turkey.d15N, shapiro.test(value[Site == "Tlajinga"])) # p = 0.4532

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.turkey.d15N, var.equal = TRUE)
# p = 0.2313
# Not reject!
```

```{r}
# "d15N_air" level in Cottontail groups
test.Cottontail.d15N <- filter(aggdata, Taxa == "Cottontail" & variable== "d15N_air")
with(test.Cottontail.d15N, shapiro.test(value[Site == "BAOX"])) # p = 0.09437
with(test.Cottontail.d15N, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2703

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Cottontail.d15N, var.equal = TRUE)
# p = 0.0321 < 0.05
# Reject!
```


# 4 Discussion

There exists an outlier point in the data set.

We can not make sure that they are same at 100% level. We can only say that at 95% confidence interval, the diets between BAOX and PATT are the same. 


# 5 Conclusion

# Citation

**Data Source**


\newpage

# Appendix

## Resource

**multiple testing**

Megan Goldman, 2008, [Statistics for Bioinformatics ](https://www.stat.berkeley.edu/~mgoldman/Section0402.pdf)
