---
title: "test"
author: "Group 2"
date: "2/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(openxlsx)
library(reshape2)
library(dplyr)
library(tidyr)
library("formattable")
library(stats)

```

## Data Cleaning
```{r, include=FALSE}
#read data
data<-read.xlsx("for_stats_peeps.xlsx")
data<-data[,-2:-6]
data<-data[,-4]
removedata<-c("MC381","MC389","MC393","MC76")
#data <- na.omit(data)
i<-1
for(i in i:4) data<-subset(data,data$ID!=removedata[i])

```

```{r, include=FALSE}
#tidy data
data2<-data[,2:8]
data2<-data2[,-3]
data2<-data2[,-5]
aggdata<-melt(data2,id=(c("Taxa","Site")))

```
Dropped 4 missing value, their IDs are MC381,MC389,MC393,MC76.

```{r,include=FALSE}
summ <- aggdata %>%                              
  group_by(Taxa,Site,variable) %>% 
  na.omit(aggdata) %>%
  summarize(min = min(value),
            median = median(value),
            mean = mean(value),
            max = max(value),
            sd = sd(value),
            count=n())
#formattable(summ)

```


```{r, echo=FALSE,warning=FALSE}
data_deer<-aggdata[which(aggdata$Taxa=="Deer"),]
data_Cottontail<-aggdata[which(aggdata$Taxa=="Cottontail"),]
data_Hare<-aggdata[which(aggdata$Taxa=="Hare"),]
data_Turkey<-aggdata[which(aggdata$Taxa=="Turkey"),]
```


## Mann Whitney U Test

Mann-Whitney U Test for deer.

```{r,echo=FALSE,warning=FALSE}
data_deer_dc<-data_deer[which(data_deer$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_deer_dc)

data_deer_apc<-data_deer[which(data_deer$variable=="ap_13C_VPDB"),] %>% na.omit(data_deer)
wilcox.test(value~Site,data=data_deer_apc)

data_deer_n<-data_deer[which(data_deer$variable=="d15N_air"),]
wilcox.test(value~Site,data=data_deer_n)
```

Mann-Whitney U Test for Cottontail.

```{r,echo=FALSE,warning=FALSE}
data_cott_dc<-data_Cottontail[which(data_Cottontail$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_cott_dc)

data_cott_apc<-data_Cottontail[which(data_Cottontail$variable=="ap_13C_VPDB"),] 
wilcox.test(value~Site,data=data_cott_apc)

data_cott_n<-data_Cottontail[which(data_Cottontail$variable=="d15N_air"),]%>% na.omit(data_Cottontail)
wilcox.test(value~Site,data=data_cott_n)
```

Mann-Whitney U Test for Hare.

```{r,echo=FALSE,warning=FALSE}
data_hare_dc<-data_Hare[which(data_Hare$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_hare_dc)

data_hare_apc<-data_Hare[which(data_Hare$variable=="ap_13C_VPDB"),] 
wilcox.test(value~Site,data=data_hare_apc)

data_hare_n<-data_Hare[which(data_Hare$variable=="d15N_air"),]
wilcox.test(value~Site,data=data_hare_n)
```

Mann-Whitney U Test for Turkey.

```{r,echo=FALSE,warning=FALSE}
data_turkey_dc<-data_Turkey[which(data_Turkey$variable=="d13C_VPDB"),]
wilcox.test(value~Site,data=data_turkey_dc)

data_turkey_apc<-data_Turkey[which(data_Turkey$variable=="ap_13C_VPDB"),] 
wilcox.test(value~Site,data=data_turkey_apc)

data_turkey_n<-data_Turkey[which(data_Turkey$variable=="d15N_air"),]
wilcox.test(value~Site,data=data_turkey_n)
```

waining mean:
 The impact of ties means the Wilcoxon rank sum distribution cannot be used to calculate exact p-values. If ties occur in our data and we have fewer than 50 observations, the wilcox.test function returns a normal approximated p-value along with a warning message that says “cannot compute exact p-value with ties”.


Since this p-value is not less than 0.05, we fail to reject the null hypothesis. 



* Mann-Whitney U Test for Cottontail.

 W=52, p-value= 0.03792

 we can reject the null hypothesis:
 two sites animals are the same


## Check Mann-Whitney U Test with theory

Because we find that Cottontail of d15N_air test reject the null hypothesis due to p-value.
We want to check Mann-Whitney U Test without R package again.

```{r,echo=FALSE,warning=FALSE}
mann <- subset(aggdata,select = -Site) %>%                         
  group_by(Taxa,variable) %>% 
  na.omit(aggdata) %>%
  summarize(rank = rank(value))
mann<- aggdata %>%                              
  group_by(Taxa,variable) %>% 
  na.omit(aggdata) %>%
  cbind(rank=mann$rank)
#formattable(mann)

mann_cott_n<-mann[which(mann$variable=="d15N_air" & mann$Taxa=="Cottontail"),]
setA<-mann_cott_n[mann_cott_n$Site=="BAOX",]
setB<-mann_cott_n[mann_cott_n$Site=="Tlajinga",]
n1<-length(setA$Taxa)
n2<-length(setA$Taxa)
R1<-sum(setA$rank)
R2<-sum(setB$rank)
U1<-n1*n2+n1*(n1+1)/2-R1
U2<-n1*n2+n2*(n2+1)/2-R2
#min(U1,U2)

```

 Alpha=0.01, test statistic(7) and our critical value(7) cannot reject

 Alpha=0.05. Since our test statistic (13) is greater than our critical value (7), we fail to reject the null hypothesis.


 Theory : https://www.statology.org/mann-whitney-u-test/
 Table : https://www.statology.org/mann-whitney-u-table/


## Next Step

We plan to try some statistic test or some regression to compare the isotopes level in two sites more precisely.

Two-Sample T-tests are as followings:
  
### "d13C_VPDB" level
  
```{r}
# Assumption
# H0 : v1  = v2
# Ha : v1 != v2
```

```{r}
# "d13C_VPDB" level in turkey groups
test.turkey.d13C <- filter(aggdata, Taxa == "Turkey" & variable== "d13C_VPDB")
with(test.turkey.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.4606
with(test.turkey.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.7027

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.turkey.d13C, var.equal = TRUE)
# p = 0.3323
# Not reject!
```

```{r}
# "d13C_VPDB" level in Cottontail groups
test.Cottontail.d13C <- filter(aggdata, Taxa == "Cottontail" & variable== "d13C_VPDB")
with(test.Cottontail.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.5969
with(test.Cottontail.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.3941

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Cottontail.d13C, var.equal = TRUE)
# p = 0.8162
# Not reject!
```

```{r}
# "d13C_VPDB" level in Deer groups
test.Deer.d13C <- filter(aggdata, Taxa == "Deer" & variable== "d13C_VPDB")
with(test.Deer.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.3914
with(test.Deer.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2177

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Deer.d13C, var.equal = TRUE)
# p = 0.6657
# Not reject!
```

```{r}
# "d13C_VPDB" level in Hare groups
test.Hare.d13C <- filter(aggdata, Taxa == "Hare" & variable== "d13C_VPDB")
with(test.Hare.d13C, shapiro.test(value[Site == "BAOX"])) # p = 0.2707
with(test.Hare.d13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2447

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.Hare.d13C, var.equal = TRUE)
# p = 0.3999
# Not reject!
```

### "d15N_air" level

```{r}
# "d15N_air" level in turkey groups
test.turkey.d15N <- filter(aggdata, Taxa == "Turkey" & variable== "d15N_air")
with(test.turkey.d15N, shapiro.test(value[Site == "BAOX"])) # p = 0.7717
with(test.turkey.d15N, shapiro.test(value[Site == "Tlajinga"])) # p = 0.4532

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.turkey.d15N, var.equal = TRUE)
# p = 0.2313
# Not reject!
```

```{r}
# "d15N_air" level in Cottontail groups
test.Cottontail.d15N <- filter(aggdata, Taxa == "Cottontail" & variable== "d15N_air")
with(test.Cottontail.d15N, shapiro.test(value[Site == "BAOX"])) # p = 0.9938
with(test.Cottontail.d15N, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2703

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Cottontail.d15N, var.equal = TRUE)
# p = 0.02954 < 0.05
# Reject!
```


```{r}
# "d15N_air" level in Deer groups
test.Deer.d15N <- filter(aggdata, Taxa == "Deer" & variable== "d15N_air")
with(test.Deer.d15N, shapiro.test(value[Site == "BAOX"])) # p = 0.5592
with(test.Deer.d15N, shapiro.test(value[Site == "Tlajinga"])) # p = 0.7696

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Deer.d15N, var.equal = TRUE)
# p = 0.3106
# Not reject!
```


```{r}
# "d15N_air" level in Hare groups
test.Hare.d15N <- filter(aggdata, Taxa == "Hare" & variable== "d15N_air")
with(test.Hare.d15N, shapiro.test(value[Site == "BAOX"])) # p = 0.3232
with(test.Hare.d15N, shapiro.test(value[Site == "Tlajinga"])) # p = 0.4432

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.Hare.d15N, var.equal = TRUE)
# p = 0.4013
# Not reject!
```


### "ap_13C_VPDB" level

```{r}
# "ap_13C_VPDB" level in turkey groups
test.turkey.ap13C <- filter(aggdata, Taxa == "Turkey" & variable== "ap_13C_VPDB")
with(test.turkey.ap13C, shapiro.test(value[Site == "BAOX"])) # p = 0.5876
with(test.turkey.ap13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.3992

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.turkey.ap13C, var.equal = TRUE)
# p = 0.2241
# Not reject!
```

```{r}
# "ap_13C_VPDB" level in Cottontail groups
test.Cottontail.ap13C <- filter(aggdata, Taxa == "Cottontail" & variable== "ap_13C_VPDB")
with(test.Cottontail.ap13C, shapiro.test(value[Site == "BAOX"])) # p = 0.1675
with(test.Cottontail.ap13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.4795

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Cottontail.ap13C, var.equal = TRUE)
# p = 0.6955
# Reject!
```


```{r}
# "ap_13C_VPDB" level in Deer groups
test.Deer.ap13C <- filter(aggdata, Taxa == "Deer" & variable== "ap_13C_VPDB")
with(test.Deer.ap13C, shapiro.test(value[Site == "BAOX"])) # p = 0.54
with(test.Deer.ap13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.3121

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.
t.test(value ~ Site, data = test.Deer.ap13C, var.equal = TRUE)
# p = 0.4406
# Not reject!
```


```{r}
# "ap_13C_VPDB" level in Hare groups
test.Hare.ap13C <- filter(aggdata, Taxa == "Hare" & variable== "ap_13C_VPDB")
with(test.Hare.ap13C, shapiro.test(value[Site == "BAOX"])) # p = 0.7399
with(test.Hare.ap13C, shapiro.test(value[Site == "Tlajinga"])) # p = 0.2754

# From the output, the two p-values are greater than the significance level 0.05 implying that the distribution of the data are not significantly different from the normal distribution. In other words, we can assume the normality.

t.test(value ~ Site, data = test.Hare.ap13C, var.equal = TRUE)
# p = 0.8815
# Not reject!
```







